---
title: "Near misses in a dynamic gambling task"
subtitle: "Studies 1 and 2 — results"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: hide
    theme: flatly
    fig_width: 10
    fig_height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center", dpi = 110)
suppressPackageStartupMessages({
  library(tidyverse); library(lme4); library(lmerTest); library(patchwork); library(knitr)
})
# run from the project root
if (basename(getwd()) == "analysis") knitr::opts_knit$set(root.dir = "..")
```

```{r helpers}
cond_levels <- c("CW", "NW", "NM", "CL")
cond_labs   <- c(CW = "clear win", NW = "narrow win", NM = "near miss", CL = "clear loss")
traj_levels <- c("catch-up", "fall-back", "stable")
cond_cols   <- c(CW = "#009E73", NW = "#56B4E9", NM = "#E69F00", CL = "#D55E00")
traj_cols   <- c("catch-up" = "#E69F00", "fall-back" = "#B25400", "stable" = "#B0982E")
dvs <- c(confidence = "Pre-race confidence", pleased = "Pleasure",
         motivation = "Motivation to continue", luck = "Luck")

# Cousineau-Morey within-subject CIs
wsci <- function(dat, group, colname, conf = 0.95) {
  g <- rlang::ensym(group)
  M <- dplyr::n_distinct(dat[[rlang::as_string(g)]]); gm <- mean(dat[[colname]])
  dat |> group_by(prolificPID) |>
    mutate(subj = mean(.data[[colname]]), yn = .data[[colname]] - subj + gm) |>
    ungroup() |> group_by(!!g) |>
    summarise(m = mean(.data[[colname]]), sdn = sd(yn), nn = n(), .groups = "drop") |>
    mutate(sem = sdn / sqrt(nn) * sqrt(M / (M - 1)),
           ci = qt(1 - (1 - conf) / 2, nn - 1) * sem, lo = m - ci, hi = m + ci)
}

# tidy a paired t-test into one row
tt_row <- function(a, b, label) {
  # note: the result object must NOT be called `t` — tibble() lets later columns
  # refer to earlier ones, so a column named `t` would shadow it mid-call.
  res <- t.test(a, b, paired = TRUE)
  tibble(contrast = label, diff = mean(a - b), t = unname(res$statistic),
         df = unname(res$parameter), p = res$p.value, dz = mean(a - b) / sd(a - b))
}
# Bold a whole row when p < .05. kable needs escape = FALSE to honour the markup.
bold_if <- function(x, hit) ifelse(hit, paste0("**", x, "**"), as.character(x))

fmt <- function(tab) {
  hit <- tab$p < .05
  tab |>
    mutate(across(c(diff, t, dz), \(x) round(x, 2)),
           p = ifelse(p < .0001, "<.0001", sprintf("%.4f", p))) |>
    mutate(across(everything(), \(col) bold_if(col, hit))) |>
    kable(escape = FALSE)
}

# ---------------------------------------------------------------------------
# Trajectory tests.
#
# One mixed model per DV: DV ~ trajectory + (1 | participant).
#
# The pre-registration named TWO PLANNED CONTRASTS — catch-up vs stable and
# catch-up vs fall-back — on motivation, pleasure and luck. Planned contrasts are
# specified before the data exist, so they are tested directly. They do not
# require a significant omnibus first, and conventionally are not corrected.
# Holm across the six is shown anyway so nothing is hidden.
#
# fall-back vs stable was NOT planned, and neither were the narrow-win contrasts
# beyond the mirror prediction in box 6. Those are gated on the omnibus (Fisher's
# protected LSD, exact for three groups) and marked when the gate fails.
# ---------------------------------------------------------------------------
traj_table <- function(trial_data, pairs) {
  ppt <- trial_data |> group_by(prolificPID, trajectory) |>
    summarise(across(all_of(names(dvs)), mean), .groups = "drop")
  out <- map_dfr(names(dvs), function(v) {
    a  <- anova(lmer(as.formula(paste(v, "~ trajectory + (1|prolificPID)")), data = trial_data))
    op <- a$`Pr(>F)`
    ww <- ppt |> select(prolificPID, trajectory, all_of(v)) |>
      pivot_wider(names_from = trajectory, values_from = all_of(v))
    map_dfr(pairs, function(pr) {
      r <- t.test(ww[[pr[[1]]]], ww[[pr[[2]]]], paired = TRUE)
      tibble(DV = dvs[[v]], omnibus_F = unname(a$`F value`), omnibus_p = op,
             contrast = pr[[3]], planned = pr[[4]],
             diff = mean(ww[[pr[[1]]]] - ww[[pr[[2]]]]), p_raw = r$p.value,
             dz = mean(ww[[pr[[1]]]] - ww[[pr[[2]]]]) / sd(ww[[pr[[1]]]] - ww[[pr[[2]]]]))
    })
  })
  # Holm within the planned family only; unplanned rows are gated on the omnibus
  out |>
    mutate(holm = ifelse(planned, p.adjust(ifelse(planned, p_raw, NA), "holm"), NA_real_),
           status = case_when(
             planned                    ~ "planned",
             !planned & omnibus_p < .05 ~ "post-hoc, omnibus passed",
             TRUE                       ~ "post-hoc, omnibus failed"),
           p = ifelse(status == "post-hoc, omnibus failed", "— not licensed —",
                      ifelse(p_raw < .0001, "<.0001", sprintf("%.4f", p_raw))),
           holm = ifelse(is.na(holm), "", sprintf("%.4f", holm)),
           dz = ifelse(status == "post-hoc, omnibus failed", NA_real_, round(dz, 2)),
           across(c(omnibus_F, diff), \(x) round(x, 2)),
           omnibus_p = ifelse(omnibus_p < .0001, "<.0001", sprintf("%.4f", omnibus_p))) |>
    select(DV, omnibus_F, omnibus_p, contrast, status, diff, dz, p, holm) |>
    # bold the rows that are both licensed and significant
    (\(d) {
      hit <- d$p != "— not licensed —" &
             suppressWarnings(ifelse(d$p == "<.0001", 0, as.numeric(d$p))) < .05
      hit[is.na(hit)] <- FALSE
      d |> mutate(across(everything(), \(col) bold_if(col, hit)))
    })()
}
```

# Notes

* Study 1: Effect of near misses in a dynamic gambling task. 
 
* Study 2: Does the shape of the near miss matter? Does the shape of the narrow win matter?

In both studies, I also collected the PGSI, just to see whether the near-miss effect is stronger in problem gamblers. Luke already warned in advance that a general Prolific panel has limited ability to test moderation, and indeed half the sample scores zero on the PGSI. So treat the PGSI results as exploratory.

Trajectory is coded from the **losing dog's** point of view, on the final part of the race: **catch-up** (gaining but ran out of track), **fall-back** (in front, passed before the line), **stable** (close, gap never changed).

Study 1's clips end on the trackside result board, Study 2's end at the finish line (I cropped the ones that showed the board because it could be a confounder). This is because Study's 1 clips were taken from a set which had the result board in view, and Study 2's clips included clips where the board was not visible. The trajectory coding is unaffected by this difference.

Ratings are 0–100 sliders. Confidence is taken **before** the race, so it serves as the check that trap assignment is not leaking; the other three are taken after.

For Study 2 (which I preregistered in aspredicted.org), I incorporated all of Luke's comments: 1. Outcome check two times in the experiment. 2. Outcome message is "Your dog did not win", rather than "unplaced". 3. The dog's names are never repeated across races. 4. The luck rating reads now “how lucky did you feel”.  

---

```{r load-s1}
s1_files <- list.files("pavlovia/data", "csv$", full.names = TRUE)
s1_files <- s1_files[!grepl("anon", s1_files)]
s1_raw <- s1_files |>
  map(\(f) read_csv(f, show_col_types = FALSE, guess_max = 100)) |>
  list_rbind() |> filter(!is.na(condition))

s1_bypid <- s1_raw |> group_by(prolificPID) |>
  summarise(n_trials = n(),
            att_n = sum(attentionCheck == TRUE, na.rm = TRUE),
            att_p = sum(attentionCheck == TRUE & attentionPass == TRUE, na.rm = TRUE),
            straight = any(c(n_distinct(motivation) == 1, n_distinct(pleased) == 1,
                             n_distinct(luck) == 1, n_distinct(confidence) == 1)),
            .groups = "drop") |>
  mutate(exclude = n_trials != 20 | att_p < att_n | straight)
s1 <- s1_raw |> filter(prolificPID %in% (s1_bypid |> filter(!exclude) |> pull(prolificPID))) |>
  mutate(condition = factor(condition, levels = cond_levels))
N1 <- n_distinct(s1$prolificPID)
s1_pm <- s1 |> group_by(prolificPID, condition) |>
  summarise(across(all_of(names(dvs)), mean), .groups = "drop")
```

# Study 1 — the near-miss effect

**N = `r N1`** of `r nrow(s1_bypid)` collected. Exclusions: incomplete
`r sum(s1_bypid$n_trials != 20)`, failed attention check
`r sum(s1_bypid$att_p < s1_bypid$att_n)`, straight-lined `r sum(s1_bypid$straight)`.

20 trials per participant, 5 of each condition.

```{r s1-traj}
# Study 1's own trajectory mapping. NOTE this is the exploratory set used at the
# time, and it differs from Study 2's locked set: it includes the floodlit night
# races (14, 22) and race 20, and it puts race 16 in catch-up. Kept as-is so the
# Study 1 panel shows what was actually looked at, not a retrofit.
s1_traj <- tibble::tribble(~raceId, ~trajectory,
   3, "catch-up",   4, "catch-up",  16, "catch-up",
   5, "fall-back",  7, "fall-back",  9, "fall-back", 15, "fall-back",
   2, "stable",    14, "stable",    20, "stable",    21, "stable", 22, "stable")
s1 <- s1 |> mutate(raceId = as.integer(raceId)) |>
  left_join(s1_traj, by = "raceId") |>
  mutate(trajectory = factor(trajectory, levels = traj_levels))
```

## Which clip served which role

Study 1 drew the trap assignment at random, so the number of times a clip served as
a near miss versus a narrow win came out uneven. Study 2 replaced this with a
rotation (see below), which removes the problem by construction.

```{r s1-clips}
s1 |> filter(condition %in% c("NW", "NM")) |>
  count(raceId, trajectory, condition) |>
  pivot_wider(names_from = condition, values_from = n, values_fill = 0) |>
  mutate(imbalance = abs(NM - NW)) |> arrange(desc(imbalance)) |>
  kable(caption = "Study 1: times each close clip served in each role (random assignment)")
```

## Ratings by outcome

```{r s1-fig1, fig.height=7}
p1 <- imap(dvs, function(label, col) {
  s <- wsci(s1_pm, condition, col)
  ggplot(s, aes(condition, m, fill = condition)) +
    geom_col(width = .66) + geom_errorbar(aes(ymin = lo, ymax = hi), width = .15) +
    scale_fill_manual(values = cond_cols, guide = "none") +
    scale_x_discrete(labels = cond_labs) + coord_cartesian(ylim = c(0, 100)) +
    labs(title = label, x = NULL, y = NULL) + theme_minimal(base_size = 11) +
    theme(axis.text.x = element_text(angle = 15, hjust = 1))
})
wrap_plots(p1, ncol = 2) +
  plot_annotation(title = sprintf("Study 1 (n=%d): mean +/- within-subject 95%% CI", N1),
                  theme = theme(plot.title = element_text(face = "bold")))
```

```{r s1-means}
s1_pm |> select(-prolificPID) |> group_by(condition) |>
  summarise(across(everything(), \(x) round(mean(x), 1))) |> kable()
```

## Key contrasts

```{r s1-h1}
w1 <- s1_pm |> select(prolificPID, condition, motivation) |>
  pivot_wider(names_from = condition, values_from = motivation)
bind_rows(
  tt_row(w1$NM, w1$CL, "near miss vs clear loss"),
  tt_row(w1$NW, w1$NM, "narrow win vs near miss"),
  tt_row(w1$CW, w1$NW, "clear win vs narrow win")
) |> fmt()
```

Confidence, rated before the outcome is known, should be flat — and is. That is the
check that trap assignment is not leaking across races.

---

```{r load-s2}
s2_files <- list.files("study2-pavlovia/data", "csv$", full.names = TRUE)
s2_files <- s2_files[!grepl("OMARTEST|anon", s2_files)]
s2_raw <- s2_files |>
  map(\(f) read_csv(f, show_col_types = FALSE, guess_max = 100,
                    col_types = cols(outcomeCheckResponse = col_character(),
                                     attentionResponse = col_character()))) |>
  list_rbind() |> filter(!is.na(condition))
stopifnot(all(s2_raw$studyId == "study2"))

s2_bypid <- s2_raw |> group_by(prolificPID) |>
  summarise(n_trials = n(),
            att_n = sum(attentionCheck == TRUE, na.rm = TRUE),
            att_p = sum(attentionCheck == TRUE & attentionPass == TRUE, na.rm = TRUE),
            out_n = sum(outcomeCheck == TRUE, na.rm = TRUE),
            out_p = sum(outcomeCheck == TRUE & outcomeCheckPass == TRUE, na.rm = TRUE),
            straight = any(c(n_distinct(motivation) == 1, n_distinct(pleased) == 1,
                             n_distinct(luck) == 1, n_distinct(confidence) == 1)),
            med_rt = median(postraceRT_ms, na.rm = TRUE), .groups = "drop") |>
  mutate(exclude = n_trials != 20 | att_p < att_n | out_p < out_n | straight | med_rt < 2000)
s2 <- s2_raw |> filter(prolificPID %in% (s2_bypid |> filter(!exclude) |> pull(prolificPID))) |>
  mutate(condition = factor(condition, levels = cond_levels),
         trajectory = factor(na_if(trajectory, ""), levels = traj_levels),
         raceId = as.integer(raceId))
N2 <- n_distinct(s2$prolificPID)
s2_pm <- s2 |> group_by(prolificPID, condition) |>
  summarise(across(all_of(names(dvs)), mean), .groups = "drop")
s2_pt <- s2 |> filter(condition == "NM") |> group_by(prolificPID, trajectory) |>
  summarise(across(all_of(names(dvs)), mean), .groups = "drop")
s2_nm <- s2 |> filter(condition == "NM", !is.na(trajectory))
```

# Study 2 — does the shape of the near miss matter?

Pre-registered on AsPredicted. **N = `r N2`** of `r nrow(s2_bypid)` collected.

## Counterbalancing

12 close clips, 4 per trajectory, plus 11 clear races. Each participant sees **2 near
misses and 2 narrow wins of every trajectory**, so the trajectory comparison is fully
within subject. Roles rotate on a participant basis rather than being drawn at
random, so across the sample every clip serves in both roles equally often, and
nobody sees a clip twice.

```{r s2-counterbalance}
s2 |> filter(condition %in% c("NW", "NM")) |>
  count(raceId, trajectory, condition) |>
  pivot_wider(names_from = condition, values_from = n, values_fill = 0) |>
  mutate(imbalance = abs(NM - NW)) |> arrange(trajectory, raceId) |>
  kable(caption = "Study 2: times each clip served in each role (rotation)")
```

## Exclusions

Study 2 added an **outcome check**: on two races, participants are asked where their
dog finished, straight after the video and before the result screen so it cannot be
read off. Much stricter than the slider check used in Study 1. This led to a significant portion of exclusions, but the remaining sample is still large enough to detect the near-miss effect. Plus it was preregistered to reach $N=100$.

```{r s2-excl}
tibble(reason = c("incomplete", "failed slider check", "failed outcome check",
                  "straight-lined", "median RT < 2 s", "TOTAL EXCLUDED", "ANALYSED"),
       n = c(sum(s2_bypid$n_trials != 20), sum(s2_bypid$att_p < s2_bypid$att_n),
             sum(s2_bypid$out_p < s2_bypid$out_n), sum(s2_bypid$straight),
             sum(s2_bypid$med_rt < 2000), sum(s2_bypid$exclude), N2)) |> kable()
```

## H1 — the near-miss effect replicates

```{r s2-fig1, fig.height=7}
p2 <- imap(dvs, function(label, col) {
  s <- wsci(s2_pm, condition, col)
  ggplot(s, aes(condition, m, fill = condition)) +
    geom_col(width = .66) + geom_errorbar(aes(ymin = lo, ymax = hi), width = .15) +
    scale_fill_manual(values = cond_cols, guide = "none") +
    scale_x_discrete(labels = cond_labs) + coord_cartesian(ylim = c(0, 100)) +
    labs(title = label, x = NULL, y = NULL) + theme_minimal(base_size = 11) +
    theme(axis.text.x = element_text(angle = 15, hjust = 1))
})
wrap_plots(p2, ncol = 2) +
  plot_annotation(title = sprintf("Study 2 (n=%d): mean +/- within-subject 95%% CI", N2),
                  theme = theme(plot.title = element_text(face = "bold")))
```

```{r s2-h1}
w2 <- s2_pm |> select(prolificPID, condition, motivation) |>
  pivot_wider(names_from = condition, values_from = motivation)
tt_row(w2$NM, w2$CL, "near miss vs clear loss (motivation)") |> fmt()
```

## H2 — trajectory

```{r s2-fig2, fig.height=7}
p3 <- imap(dvs, function(label, col) {
  s <- wsci(s2_pt, trajectory, col)
  ggplot(s, aes(trajectory, m, fill = trajectory)) +
    geom_col(width = .6) + geom_errorbar(aes(ymin = lo, ymax = hi), width = .14) +
    scale_fill_manual(values = traj_cols, guide = "none") +
    coord_cartesian(ylim = c(0, 100)) +
    labs(title = label, x = NULL, y = NULL) + theme_minimal(base_size = 11) +
    theme(axis.text.x = element_text(angle = 12, hjust = 1))
})
wrap_plots(p3, ncol = 2) +
  plot_annotation(title = sprintf("Study 2 (n=%d): NEAR MISS trials only, by trajectory", N2),
                  theme = theme(plot.title = element_text(face = "bold")))
```

**catch-up vs stable** and **catch-up vs fall-back** were named as planned contrasts
in the pre-registration, so they are tested directly — a planned contrast does not
need a significant omnibus first. `holm` is Holm across the six planned contrasts
(2 contrasts x 3 DVs).

```{r s2-traj-tests}
traj_table(s2_nm, list(
  list("catch-up",  "fall-back", "catch-up vs fall-back", TRUE),
  list("catch-up",  "stable",    "catch-up vs stable",    TRUE),
  list("fall-back", "stable",    "fall-back vs stable",   FALSE))) |> kable(escape = FALSE)
```

**The result.** Motivation is flat across trajectory. Pleasure is not: a **catch-up**
near miss feels distinctly less bad than either of the others, and fall-back and
stable are indistinguishable from each other. So the shape of a near miss changes how
it *feels* without changing how much people want to carry on — affect and drive come
apart.

Confidence stays flat, as it must.

## Within-clip view

Because the same clip is a narrow win for some participants and a near miss for
others, subtracting one from the other **within a clip** removes everything about that
particular race.

```{r s2-fig3, fig.height=5}
clip <- s2 |> filter(condition %in% c("NW", "NM"), !is.na(trajectory)) |>
  group_by(raceId, trajectory, condition) |>
  summarise(across(all_of(names(dvs)), mean), .groups = "drop") |>
  pivot_longer(all_of(names(dvs)), names_to = "col", values_to = "val") |>
  pivot_wider(names_from = condition, values_from = val) |>
  mutate(delta = NM - NW,
         dv = factor(col, levels = names(dvs), labels = unname(dvs)))
cm <- clip |> group_by(dv, trajectory) |>
  summarise(m = mean(delta), se = sd(delta) / sqrt(n()), .groups = "drop")
ggplot(clip, aes(trajectory, delta, colour = trajectory)) +
  geom_hline(yintercept = 0, colour = "grey40") +
  geom_point(size = 2.6, alpha = .75, position = position_jitter(width = .07, seed = 1)) +
  geom_errorbar(data = cm, inherit.aes = FALSE,
                aes(trajectory, ymin = m - se, ymax = m + se), width = .13) +
  geom_point(data = cm, inherit.aes = FALSE, aes(trajectory, m), shape = 95, size = 12) +
  facet_wrap(~dv, scales = "free_y") +
  scale_colour_manual(values = traj_cols, guide = "none") +
  labs(title = "Within-clip: near miss minus narrow win, same footage",
       subtitle = "each point is one race; below zero = the near miss cost more",
       x = NULL, y = "NM - NW (rating points)") +
  theme_minimal(base_size = 11) + theme(plot.title = element_text(face = "bold"))
```

## The winner's side of the same clips

The same clip seen from the winner's trap. A fall-back clip means the winner came
from behind; a catch-up clip means the winner led and was nearly caught.

Pre-registered as another research question: does coming from behind to win beat
holding on?

```{r nw-plot-fn}
nw_panels <- function(dat, N, title) {
  pw <- dat |> filter(condition == "NW", !is.na(trajectory)) |>
    group_by(prolificPID, trajectory) |>
    summarise(across(all_of(names(dvs)), mean), .groups = "drop")
  ps <- imap(dvs, function(label, col) {
    s <- wsci(pw, trajectory, col)
    ggplot(s, aes(trajectory, m, fill = trajectory)) +
      geom_col(width = .6) + geom_errorbar(aes(ymin = lo, ymax = hi), width = .14) +
      scale_fill_manual(values = traj_cols, guide = "none") +
      coord_cartesian(ylim = c(0, 100)) +
      labs(title = label, x = NULL, y = NULL) + theme_minimal(base_size = 11) +
      theme(axis.text.x = element_text(angle = 12, hjust = 1))
  })
  wrap_plots(ps, ncol = 2) +
    plot_annotation(title = sprintf("%s (n=%d): NARROW WIN trials only, by clip trajectory", title, N),
                    theme = theme(plot.title = element_text(face = "bold")))
}
```

```{r s2-nw-fig, fig.height=7}
nw_panels(s2, N2, "Study 2")
```

Box 6 of the pre-registration named one comparison here — *coming from behind to win
vs holding on* — as an exploratory but pre-specified prediction, so it is tested
directly. The other two were not specified and are gated on the omnibus.

```{r s2-nw-tests}
s2_nwtr <- s2 |> filter(condition == "NW", !is.na(trajectory))
traj_table(s2_nwtr, list(
  list("fall-back", "catch-up", "came from behind vs held on",        TRUE),
  list("fall-back", "stable",   "came from behind vs never in doubt", FALSE),
  list("catch-up",  "stable",   "held on vs never in doubt",          FALSE))) |> kable(escape = FALSE)
```

**Pleasure is at ceiling** — 91.4, 92.0, 91.5 out of 100. Every win feels great, so
there is no room to detect a difference. Read the flat pleasure panel as a ceiling
effect; I don't think we can conclude no difference.

**Luck** does move, and its omnibus passes (F = 7.45, p = .0007): both dramatic wins
feel luckier than a comfortable one.

**Motivation.** The pre-specified prediction is supported: coming from behind to win
(78.2) beats holding on (75.4), p = .017, dz = 0.24. Because it was named in advance
it is tested directly, without needing the omnibus — which is just as well, since the
omnibus sits at p = .068. Worth being straight about the strength of it: it is a
pre-specified but *exploratory* prediction, the effect is small, and it would not
survive a blanket correction across every contrast in this report. Treat it as
encouraging rather than established.

### The asymmetry

```{r asym}
tibble(
  role = c("loser (near miss)", "winner (narrow win)"),
  `moves with trajectory` = c("pleasure", "luck, motivation (small)"),
  `does not` = c("motivation, luck", "—"),
  `cannot tell` = c("—", "pleasure (at ceiling)")
) |> kable()
```

Losing: the shape changes how bad it feels, but not whether you carry on.
Winning: the shape changes how lucky you felt, and moves motivation a little in the
predicted direction. How good it feels cannot be assessed either way, because every
win is rated around 92.

### Study 1, same split

Shown for comparison only. Study 1 was exploratory in this regard, its trajectory labels were partly
reconstructed after the fact, and its stimulus set includes clips later dropped from
Study 2.

```{r s1-nw-fig, fig.height=7}
nw_panels(s1, N1, "Study 1")
```

```{r s1-nm-nw-compare}
bind_rows(
  s1 |> filter(condition %in% c("NM", "NW"), !is.na(trajectory)) |>
    group_by(prolificPID, condition, trajectory) |>
    summarise(across(all_of(names(dvs)), mean), .groups = "drop") |>
    group_by(condition, trajectory) |>
    summarise(across(all_of(names(dvs)), \(x) round(mean(x), 1)), .groups = "drop") |>
    mutate(study = "Study 1", .before = 1),
  s2 |> filter(condition %in% c("NM", "NW"), !is.na(trajectory)) |>
    group_by(prolificPID, condition, trajectory) |>
    summarise(across(all_of(names(dvs)), mean), .groups = "drop") |>
    group_by(condition, trajectory) |>
    summarise(across(all_of(names(dvs)), \(x) round(mean(x), 1)), .groups = "drop") |>
    mutate(study = "Study 2", .before = 1)
) |> arrange(condition, trajectory, study) |>
  kable(caption = "Both studies, both roles, by trajectory")
```

## PGSI (exploratory)

Pre-registered as exploratory. Half the sample scores zero on the PGSI, so a general
Prolific panel has limited ability to test moderation (as noted by Luke before running the task on Prolific).

```{r s2-pgsi-dist}
pg <- s2 |> group_by(prolificPID) |> summarise(pgsi = first(pgsi_total), .groups = "drop") |>
  mutate(band = if_else(pgsi >= 3, "elevated (PGSI 3+)", "low (PGSI 0-2)"))
table(cut(pg$pgsi, c(-1, 0, 2, 7, 27), labels = c("0", "1-2", "3-7", "8+"))) |>
  as.data.frame() |> setNames(c("PGSI band", "n")) |> kable()
```

```{r s2-pgsi-fig, fig.height=7}
ptb <- s2_pt |> left_join(pg, by = "prolificPID")
band_cols <- c("elevated (PGSI 3+)" = "#CC3311", "low (PGSI 0-2)" = "#4477AA")
p4 <- imap(dvs, function(label, col) {
  s <- ptb |> group_by(band) |> group_modify(~ wsci(.x, trajectory, col)) |> ungroup()
  ggplot(s, aes(trajectory, m, colour = band, group = band)) +
    geom_line(linewidth = .8) + geom_errorbar(aes(ymin = lo, ymax = hi), width = .12) +
    geom_point(size = 2.5) + scale_colour_manual(values = band_cols) +
    coord_cartesian(ylim = c(0, 100)) +
    labs(title = label, x = NULL, y = NULL, colour = NULL) +
    theme_minimal(base_size = 11) + theme(axis.text.x = element_text(angle = 12, hjust = 1))
})
wrap_plots(p4, ncol = 2, guides = "collect") +
  plot_annotation(title = "Near-miss trials by trajectory, split by PGSI",
                  theme = theme(plot.title = element_text(face = "bold"))) &
  theme(legend.position = "bottom")
```

```{r s2-pgsi-tests}
pp <- s2_nm |> left_join(pg, by = "prolificPID") |>
  group_by(prolificPID, band, pgsi, trajectory) |>
  summarise(pleased = mean(pleased), .groups = "drop") |>
  pivot_wider(names_from = trajectory, values_from = pleased)
names(pp)[4:6] <- c("cu", "fb", "st"); pp <- pp |> mutate(cu_fb = cu - fb)
map_dfr(unique(pp$band), function(b) {
  s <- pp |> filter(band == b)
  tt_row(s$cu, s$fb, paste0("catch-up vs fall-back — ", b)) |> mutate(n = nrow(s), .after = 1)
}) |> fmt()
```

```{r s2-pgsi-interaction}
tibble(
  specification = c("PGSI 3+ vs 0-2 (standard cutoff)", "continuous PGSI",
                    "log(1+PGSI)", "Spearman rank"),
  p = c(t.test(cu_fb ~ band, data = pp)$p.value,
        anova(lmer(pleased ~ trajectory * scale(pgsi) + (1|prolificPID),
                   data = s2_nm |> left_join(pg, by = "prolificPID")))["trajectory:scale(pgsi)", "Pr(>F)"],
        anova(lmer(pleased ~ trajectory * scale(log1p(pgsi)) + (1|prolificPID),
                   data = s2_nm |> left_join(pg, by = "prolificPID")))["trajectory:scale(log1p(pgsi))", "Pr(>F)"],
        cor.test(pp$pgsi, pp$cu_fb, method = "spearman", exact = FALSE)$p.value)
) |> mutate(p = sprintf("%.3f", p)) |>
  kable(caption = "Interaction p depends on how PGSI is modelled — treat as suggestive only")
```

TThe catch-up advantage is about three times larger in the elevated group. The
interaction reaches significance when the sample is split at the standard PGSI
cutoff of 3 (moderate risk or above), but not when PGSI is entered as a continuous
term.

---

# Summary

| | Study 1 | Study 2 |
|---|---|---|
| N analysed | `r N1` | `r N2` |
| pre-registered | no | yes |
| outcome check | no | yes |
| near-miss effect (NM vs CL, motivation) | `r sprintf("+%.1f, p = %.1g", mean(w1$NM - w1$CL), t.test(w1$NM, w1$CL, paired = TRUE)$p.value)` | `r sprintf("+%.1f, p = %.1g", mean(w2$NM - w2$CL), t.test(w2$NM, w2$CL, paired = TRUE)$p.value)` |
| trajectory on motivation | — | null |
| trajectory on pleasure | — | catch-up > both others |

**What we can say.** Near misses keep people playing more than clear losses, in a
naturalistic dynamic task, replicated and pre-registered. The shape of the near miss
does not change that — but it does change how bad the loss feels, with a catch-up
near miss the least unpleasant of the three.

**What we cannot say yet.** Whether self-reported motivation predicts actual
persistence, and whether the pleasure effect is stronger in problem gamblers. Both
need studies designed for the purpose. 

**Possible follow-ups.** 1. A study that measures actual persistence, not just self-reported motivation. 2. A study that recruits a sample of problem gamblers to test whether the near-miss effect is stronger in that population. 3. The cashout manipulation study. Here we can ask a) whether the NM effect is affected by the cashout decision (or the possibility of it), and b) whether the trajectory modulates that effect. Additionally, we can c) manipulate the amount of cashout depending on the trajectory, to see whether, for example, catching up leads to less cashouts or lower probability of accepting a given cashout amount, etc. 

```{r session}
sessionInfo()
```
