---
title: "Near misses in a dynamic gambling task"
subtitle: "Studies 1 and 2 — analysis report"
author: "Omar D. Perez, Luke Clark, Steve Sharman, Mike Aitken"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: hide
    theme: flatly
    fig_width: 10
    fig_height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center", dpi = 110)
suppressPackageStartupMessages({
  library(tidyverse); library(lme4); library(lmerTest); library(patchwork); library(knitr)
})
# run from the project root
if (basename(getwd()) == "analysis") knitr::opts_knit$set(root.dir = "..")
```

```{r helpers}
cond_levels <- c("CW", "NW", "NM", "CL")
cond_labs   <- c(CW = "clear win", NW = "narrow win", NM = "near miss", CL = "clear loss")
traj_levels <- c("catch-up", "fall-back", "stable")
cond_cols   <- c(CW = "#009E73", NW = "#56B4E9", NM = "#E69F00", CL = "#D55E00")
traj_cols   <- c("catch-up" = "#E69F00", "fall-back" = "#B25400", "stable" = "#B0982E")
dvs <- c(confidence = "Pre-race confidence", pleased = "Pleasure",
         motivation = "Motivation to continue", luck = "Luck")

# Cousineau-Morey within-subject CIs
wsci <- function(dat, group, colname, conf = 0.95) {
  g <- rlang::ensym(group)
  M <- dplyr::n_distinct(dat[[rlang::as_string(g)]]); gm <- mean(dat[[colname]])
  dat |> group_by(prolificPID) |>
    mutate(subj = mean(.data[[colname]]), yn = .data[[colname]] - subj + gm) |>
    ungroup() |> group_by(!!g) |>
    summarise(m = mean(.data[[colname]]), sdn = sd(yn), nn = n(), .groups = "drop") |>
    mutate(sem = sdn / sqrt(nn) * sqrt(M / (M - 1)),
           ci = qt(1 - (1 - conf) / 2, nn - 1) * sem, lo = m - ci, hi = m + ci)
}

# tidy a paired t-test into one row
tt_row <- function(a, b, label) {
  # note: the result object must NOT be called `t` — tibble() lets later columns
  # refer to earlier ones, so a column named `t` would shadow it mid-call.
  res <- t.test(a, b, paired = TRUE)
  tibble(contrast = label, diff = mean(a - b), t = unname(res$statistic),
         df = unname(res$parameter), p = res$p.value, dz = mean(a - b) / sd(a - b))
}
fmt <- function(tab) tab |>
  mutate(across(c(diff, t, dz), \(x) round(x, 2)),
         p = ifelse(p < .0001, "<.0001", sprintf("%.4f", p))) |> kable()
```

# What this is

Two experiments asking whether **near misses** — almost winning — keep people
gambling, using real greyhound race footage rather than a slot machine.

Participants back a dog, watch the race, and find out where it finished. What they
don't know is that the trap they are assigned to determines the outcome. **The same
clip is therefore a narrow win for some participants and a near miss for others**,
which means every race acts as its own control.

| condition | clip | assigned trap | they see |
|---|---|---|---|
| **CW** clear win | clear race | winner | finishes 1st |
| **NW** narrow win | close race | winner | finishes 1st |
| **NM** near miss | close race | runner-up | finishes 2nd |
| **CL** clear loss | clear race | a back trap | did not win |

**Study 1** (exploratory) tested the basic near-miss effect. **Study 2**
(pre-registered) replicated it and asked a further question: does the *shape* of the
near miss matter?

Trajectory is judged from the **losing dog's** point of view, on the final part of
the race:

- **catch-up** — it was gaining on the leader but the race ended first
- **fall-back** — it was in front and was passed before the line
- **stable** — it finished close but the gap never changed

> **The two studies are never pooled.** Different samples, and different stimuli —
> Study 1's clips end on the trackside result board, Study 2's end at the finish line.

---

```{r load-s1}
s1_files <- list.files("pavlovia/data", "csv$", full.names = TRUE)
s1_files <- s1_files[!grepl("anon", s1_files)]
s1_raw <- s1_files |>
  map(\(f) read_csv(f, show_col_types = FALSE, guess_max = 100)) |>
  list_rbind() |> filter(!is.na(condition))

s1_bypid <- s1_raw |> group_by(prolificPID) |>
  summarise(n_trials = n(),
            att_n = sum(attentionCheck == TRUE, na.rm = TRUE),
            att_p = sum(attentionCheck == TRUE & attentionPass == TRUE, na.rm = TRUE),
            straight = any(c(n_distinct(motivation) == 1, n_distinct(pleased) == 1,
                             n_distinct(luck) == 1, n_distinct(confidence) == 1)),
            .groups = "drop") |>
  mutate(exclude = n_trials != 20 | att_p < att_n | straight)
s1 <- s1_raw |> filter(prolificPID %in% (s1_bypid |> filter(!exclude) |> pull(prolificPID))) |>
  mutate(condition = factor(condition, levels = cond_levels))
N1 <- n_distinct(s1$prolificPID)
s1_pm <- s1 |> group_by(prolificPID, condition) |>
  summarise(across(all_of(names(dvs)), mean), .groups = "drop")
```

# Study 1 — the near-miss effect

**N = `r N1`** of `r nrow(s1_bypid)` collected. Exclusions: incomplete
`r sum(s1_bypid$n_trials != 20)`, failed attention check
`r sum(s1_bypid$att_p < s1_bypid$att_n)`, straight-lined `r sum(s1_bypid$straight)`.

20 trials per participant, 5 of each condition. Ratings are 0–100 sliders; confidence
is taken **before** the race, the other three after.

## Ratings by outcome

```{r s1-fig1, fig.height=7}
p1 <- imap(dvs, function(label, col) {
  s <- wsci(s1_pm, condition, col)
  ggplot(s, aes(condition, m, fill = condition)) +
    geom_col(width = .66) + geom_errorbar(aes(ymin = lo, ymax = hi), width = .15) +
    scale_fill_manual(values = cond_cols, guide = "none") +
    scale_x_discrete(labels = cond_labs) + coord_cartesian(ylim = c(0, 100)) +
    labs(title = label, x = NULL, y = NULL) + theme_minimal(base_size = 11) +
    theme(axis.text.x = element_text(angle = 15, hjust = 1))
})
wrap_plots(p1, ncol = 2) +
  plot_annotation(title = sprintf("Study 1 (n=%d): mean +/- within-subject 95%% CI", N1),
                  theme = theme(plot.title = element_text(face = "bold")))
```

```{r s1-means}
s1_pm |> select(-prolificPID) |> group_by(condition) |>
  summarise(across(everything(), \(x) round(mean(x), 1))) |> kable()
```

## The key contrast

Both a near miss and a clear loss are losses. The question is whether *almost*
winning keeps people playing more than clearly losing.

```{r s1-h1}
w1 <- s1_pm |> select(prolificPID, condition, motivation) |>
  pivot_wider(names_from = condition, values_from = motivation)
bind_rows(
  tt_row(w1$NM, w1$CL, "near miss vs clear loss"),
  tt_row(w1$NW, w1$NM, "narrow win vs near miss"),
  tt_row(w1$CW, w1$NW, "clear win vs narrow win")
) |> fmt()
```

Confidence, rated before the outcome is known, should be flat — and is. That is the
check that trap assignment is not leaking.

---

```{r load-s2}
s2_files <- list.files("study2-pavlovia/data", "csv$", full.names = TRUE)
s2_files <- s2_files[!grepl("OMARTEST|anon", s2_files)]
s2_raw <- s2_files |>
  map(\(f) read_csv(f, show_col_types = FALSE, guess_max = 100,
                    col_types = cols(outcomeCheckResponse = col_character(),
                                     attentionResponse = col_character()))) |>
  list_rbind() |> filter(!is.na(condition))
stopifnot(all(s2_raw$studyId == "study2"))

s2_bypid <- s2_raw |> group_by(prolificPID) |>
  summarise(n_trials = n(),
            att_n = sum(attentionCheck == TRUE, na.rm = TRUE),
            att_p = sum(attentionCheck == TRUE & attentionPass == TRUE, na.rm = TRUE),
            out_n = sum(outcomeCheck == TRUE, na.rm = TRUE),
            out_p = sum(outcomeCheck == TRUE & outcomeCheckPass == TRUE, na.rm = TRUE),
            straight = any(c(n_distinct(motivation) == 1, n_distinct(pleased) == 1,
                             n_distinct(luck) == 1, n_distinct(confidence) == 1)),
            med_rt = median(postraceRT_ms, na.rm = TRUE), .groups = "drop") |>
  mutate(exclude = n_trials != 20 | att_p < att_n | out_p < out_n | straight | med_rt < 2000)
s2 <- s2_raw |> filter(prolificPID %in% (s2_bypid |> filter(!exclude) |> pull(prolificPID))) |>
  mutate(condition = factor(condition, levels = cond_levels),
         trajectory = factor(na_if(trajectory, ""), levels = traj_levels),
         raceId = as.integer(raceId))
N2 <- n_distinct(s2$prolificPID)
s2_pm <- s2 |> group_by(prolificPID, condition) |>
  summarise(across(all_of(names(dvs)), mean), .groups = "drop")
s2_pt <- s2 |> filter(condition == "NM") |> group_by(prolificPID, trajectory) |>
  summarise(across(all_of(names(dvs)), mean), .groups = "drop")
s2_nm <- s2 |> filter(condition == "NM", !is.na(trajectory))
```

# Study 2 — does the shape of the near miss matter?

Pre-registered on AsPredicted. **N = `r N2`** of `r nrow(s2_bypid)` collected.

## Design

12 close clips, 4 per trajectory, plus 11 clear races. Each participant sees **2 near
misses and 2 narrow wins of every trajectory**, so the trajectory comparison is fully
within subject.

Roles rotate on a participant index rather than being drawn at random, so across the
sample every clip serves as narrow win and as near miss equally often. Nobody sees a
clip twice.

```{r s2-counterbalance}
s2 |> filter(condition %in% c("NW", "NM")) |>
  count(raceId, trajectory, condition) |>
  pivot_wider(names_from = condition, values_from = n, values_fill = 0) |>
  arrange(trajectory, raceId) |> kable(caption = "Times each clip served in each role")
```

## Exclusions

Study 2 added an **outcome check**: on two races, participants are asked where their
dog finished, immediately after the video and before the result screen, so the answer
cannot be read off it. It is much stricter than the slider check used in Study 1.

```{r s2-excl}
tibble(reason = c("incomplete", "failed slider check", "failed outcome check",
                  "straight-lined", "median RT < 2 s", "TOTAL EXCLUDED", "ANALYSED"),
       n = c(sum(s2_bypid$n_trials != 20), sum(s2_bypid$att_p < s2_bypid$att_n),
             sum(s2_bypid$out_p < s2_bypid$out_n), sum(s2_bypid$straight),
             sum(s2_bypid$med_rt < 2000), sum(s2_bypid$exclude), N2)) |> kable()
```

## H1 — the near-miss effect replicates

```{r s2-fig1, fig.height=7}
p2 <- imap(dvs, function(label, col) {
  s <- wsci(s2_pm, condition, col)
  ggplot(s, aes(condition, m, fill = condition)) +
    geom_col(width = .66) + geom_errorbar(aes(ymin = lo, ymax = hi), width = .15) +
    scale_fill_manual(values = cond_cols, guide = "none") +
    scale_x_discrete(labels = cond_labs) + coord_cartesian(ylim = c(0, 100)) +
    labs(title = label, x = NULL, y = NULL) + theme_minimal(base_size = 11) +
    theme(axis.text.x = element_text(angle = 15, hjust = 1))
})
wrap_plots(p2, ncol = 2) +
  plot_annotation(title = sprintf("Study 2 (n=%d): mean +/- within-subject 95%% CI", N2),
                  theme = theme(plot.title = element_text(face = "bold")))
```

```{r s2-h1}
w2 <- s2_pm |> select(prolificPID, condition, motivation) |>
  pivot_wider(names_from = condition, values_from = motivation)
tt_row(w2$NM, w2$CL, "near miss vs clear loss (motivation)") |> fmt()
```

## H2 — trajectory

```{r s2-fig2, fig.height=7}
p3 <- imap(dvs, function(label, col) {
  s <- wsci(s2_pt, trajectory, col)
  ggplot(s, aes(trajectory, m, fill = trajectory)) +
    geom_col(width = .6) + geom_errorbar(aes(ymin = lo, ymax = hi), width = .14) +
    scale_fill_manual(values = traj_cols, guide = "none") +
    coord_cartesian(ylim = c(0, 100)) +
    labs(title = label, x = NULL, y = NULL) + theme_minimal(base_size = 11) +
    theme(axis.text.x = element_text(angle = 12, hjust = 1))
})
wrap_plots(p3, ncol = 2) +
  plot_annotation(title = sprintf("Study 2 (n=%d): NEAR MISS trials only, by trajectory", N2),
                  theme = theme(plot.title = element_text(face = "bold")))
```

```{r s2-traj-tests}
map_dfr(names(dvs), function(v) {
  ww <- s2_pt |> select(prolificPID, trajectory, all_of(v)) |>
    pivot_wider(names_from = trajectory, values_from = all_of(v))
  a <- anova(lmer(as.formula(paste(v, "~ trajectory + (1|prolificPID)")), data = s2_nm))
  bind_rows(
    tt_row(ww$`catch-up`, ww$`fall-back`, "catch-up vs fall-back"),
    tt_row(ww$`catch-up`, ww$stable,      "catch-up vs stable"),
    tt_row(ww$`fall-back`, ww$stable,     "fall-back vs stable")
  ) |> mutate(DV = dvs[[v]], anova_p = a$`Pr(>F)`, .before = 1)
}) |> mutate(across(c(diff, t, dz), \(x) round(x, 2)),
             p = ifelse(p < .0001, "<.0001", sprintf("%.4f", p)),
             anova_p = ifelse(anova_p < .0001, "<.0001", sprintf("%.4f", anova_p))) |> kable()
```

**The result.** Motivation is flat across trajectory. Pleasure is not: a **catch-up**
near miss feels distinctly less bad than either of the others, and fall-back and
stable are indistinguishable from each other. So the shape of a near miss changes how
it *feels* without changing how much people want to carry on — affect and drive come
apart.

Confidence stays flat, as it must.

## Within-clip view

Because the same clip is a narrow win for some participants and a near miss for
others, subtracting one from the other **within a clip** removes everything about that
particular race.

```{r s2-fig3, fig.height=5}
clip <- s2 |> filter(condition %in% c("NW", "NM"), !is.na(trajectory)) |>
  group_by(raceId, trajectory, condition) |>
  summarise(across(all_of(names(dvs)), mean), .groups = "drop") |>
  pivot_longer(all_of(names(dvs)), names_to = "col", values_to = "val") |>
  pivot_wider(names_from = condition, values_from = val) |>
  mutate(delta = NM - NW,
         dv = factor(col, levels = names(dvs), labels = unname(dvs)))
cm <- clip |> group_by(dv, trajectory) |>
  summarise(m = mean(delta), se = sd(delta) / sqrt(n()), .groups = "drop")
ggplot(clip, aes(trajectory, delta, colour = trajectory)) +
  geom_hline(yintercept = 0, colour = "grey40") +
  geom_point(size = 2.6, alpha = .75, position = position_jitter(width = .07, seed = 1)) +
  geom_errorbar(data = cm, inherit.aes = FALSE,
                aes(trajectory, ymin = m - se, ymax = m + se), width = .13) +
  geom_point(data = cm, inherit.aes = FALSE, aes(trajectory, m), shape = 95, size = 12) +
  facet_wrap(~dv, scales = "free_y") +
  scale_colour_manual(values = traj_cols, guide = "none") +
  labs(title = "Within-clip: near miss minus narrow win, same footage",
       subtitle = "each point is one race; below zero = the near miss cost more",
       x = NULL, y = "NM - NW (rating points)") +
  theme_minimal(base_size = 11) + theme(plot.title = element_text(face = "bold"))
```

## PGSI (exploratory)

Pre-registered as exploratory. Half the sample scores zero on the PGSI, so a general
Prolific panel has limited ability to test moderation.

```{r s2-pgsi-dist}
pg <- s2 |> group_by(prolificPID) |> summarise(pgsi = first(pgsi_total), .groups = "drop") |>
  mutate(band = if_else(pgsi >= 3, "elevated (PGSI 3+)", "low (PGSI 0-2)"))
table(cut(pg$pgsi, c(-1, 0, 2, 7, 27), labels = c("0", "1-2", "3-7", "8+"))) |>
  as.data.frame() |> setNames(c("PGSI band", "n")) |> kable()
```

```{r s2-pgsi-fig, fig.height=7}
ptb <- s2_pt |> left_join(pg, by = "prolificPID")
band_cols <- c("elevated (PGSI 3+)" = "#CC3311", "low (PGSI 0-2)" = "#4477AA")
p4 <- imap(dvs, function(label, col) {
  s <- ptb |> group_by(band) |> group_modify(~ wsci(.x, trajectory, col)) |> ungroup()
  ggplot(s, aes(trajectory, m, colour = band, group = band)) +
    geom_line(linewidth = .8) + geom_errorbar(aes(ymin = lo, ymax = hi), width = .12) +
    geom_point(size = 2.5) + scale_colour_manual(values = band_cols) +
    coord_cartesian(ylim = c(0, 100)) +
    labs(title = label, x = NULL, y = NULL, colour = NULL) +
    theme_minimal(base_size = 11) + theme(axis.text.x = element_text(angle = 12, hjust = 1))
})
wrap_plots(p4, ncol = 2, guides = "collect") +
  plot_annotation(title = "Near-miss trials by trajectory, split by PGSI",
                  theme = theme(plot.title = element_text(face = "bold"))) &
  theme(legend.position = "bottom")
```

```{r s2-pgsi-tests}
pp <- s2_nm |> left_join(pg, by = "prolificPID") |>
  group_by(prolificPID, band, pgsi, trajectory) |>
  summarise(pleased = mean(pleased), .groups = "drop") |>
  pivot_wider(names_from = trajectory, values_from = pleased)
names(pp)[4:6] <- c("cu", "fb", "st"); pp <- pp |> mutate(cu_fb = cu - fb)
map_dfr(unique(pp$band), function(b) {
  s <- pp |> filter(band == b)
  tt_row(s$cu, s$fb, paste0("catch-up vs fall-back — ", b)) |> mutate(n = nrow(s), .after = 1)
}) |> fmt()
```

```{r s2-pgsi-interaction}
tibble(
  specification = c("median split (PGSI 3+ vs 0-2)", "continuous PGSI",
                    "log(1+PGSI)", "Spearman rank"),
  p = c(t.test(cu_fb ~ band, data = pp)$p.value,
        anova(lmer(pleased ~ trajectory * scale(pgsi) + (1|prolificPID),
                   data = s2_nm |> left_join(pg, by = "prolificPID")))["trajectory:scale(pgsi)", "Pr(>F)"],
        anova(lmer(pleased ~ trajectory * scale(log1p(pgsi)) + (1|prolificPID),
                   data = s2_nm |> left_join(pg, by = "prolificPID")))["trajectory:scale(log1p(pgsi))", "Pr(>F)"],
        cor.test(pp$pgsi, pp$cu_fb, method = "spearman", exact = FALSE)$p.value)
) |> mutate(p = sprintf("%.3f", p)) |>
  kable(caption = "Interaction p depends on how PGSI is modelled — treat as suggestive only")
```

The catch-up advantage is about three times larger in the elevated group, but the
interaction is significant on a median split and not as a linear term. **It depends on
the analytic choice, so it is a hypothesis for a future study rather than a result.**

---

# Summary

| | Study 1 | Study 2 |
|---|---|---|
| N analysed | `r N1` | `r N2` |
| pre-registered | no | yes |
| outcome check | no | yes |
| near-miss effect (NM vs CL, motivation) | `r sprintf("+%.1f, p = %.1g", mean(w1$NM - w1$CL), t.test(w1$NM, w1$CL, paired = TRUE)$p.value)` | `r sprintf("+%.1f, p = %.1g", mean(w2$NM - w2$CL), t.test(w2$NM, w2$CL, paired = TRUE)$p.value)` |
| trajectory on motivation | — | null |
| trajectory on pleasure | — | catch-up > both others |

**What we can say.** Near misses keep people playing more than clear losses, in a
naturalistic dynamic task, replicated and pre-registered. The shape of the near miss
does not change that — but it does change how bad the loss feels, with a catch-up
near miss the least unpleasant of the three.

**What we cannot say yet.** Whether self-reported motivation predicts actual
persistence, and whether the pleasure effect is stronger in problem gamblers. Both
need studies designed for the purpose.

```{r session}
sessionInfo()
```
